# *Reproducible Research: Peer Assesment 1*
#### *Denizhan EDEER*

You can find relevant Readme.Rmd file at [here](https://github.com/denizhanedeer/RepData_PeerAssessment1) for more information about the project.

###Load and Process Data

Activity data is downloaded and set at current working directory. You can also reach data from [here.](https://github.com/denizhanedeer/RepData_PeerAssessment1)

```{r}
activity12 <- read.csv("activity.csv",header = TRUE)
```

After we obtain data we need to figure out what is inside of it and do we need to reorginize or process anything. So we can use str(), head(), tail() built-in functions. 

````{r}
head(activity12)
tail(activity12)
str(activity12)
```

As we see we have **NA's** at observations. It is nice to know how important and how much percent of data is missing. 
````{r}
mean(is.na(activity12$steps))
````

So we know %13 percent of data missing and we would like to remove them. We have many options to achieve that with R but I would like to show complete.cases() function which removes all row from the data frame. So we can analyze only the days that we have "steps" observation. 
````{r}
activity <- activity12[complete.cases(activity12),]
mean(is.na(activity))
````
Second step is dates. From the str() function result we check date is Factor. So we can change it to date format by as.Date() function. 
````{r}
activity$date <- as.Date(activity$date)
str(activity$date)
```

###Mean Total Number of Steps Per Day

Our data is removed from missing values so we only need to iterate of the dates to find sum() or mean() of the steps. We can acheive it with tapply function or aggregate function from stats library.
````{r}
library(stats)
total_steps <- aggregate(activity$steps,by=list(activity$date),FUN=sum)
## it is nice to set col names after aggregation. 
names(total_steps) <- c("date","sum_of_steps")
## now we take mean of steps
mean_steps <- aggregate(activity$steps,by=list(activity$date),FUN=mean)
names(mean_steps) <- c("date","mean_of_steps")
````
Just check what we have
````{r}
head(total_steps)
head(mean_steps)
````
Time to show total number of steps per day as diagram:
````{r,fig.height=6}
hist(total_steps$sum_of_steps,breaks=20,col="blue",xlab="Total Steps",main="Total Number of Steps Per Day\n NA removed")
abline(v=mean(total_steps$sum_of_steps),lwd=7,col="yellow")
````
Mean of steps per day with median value as vertical line
````{r,fig.height=6}
hist(mean_steps$mean_of_steps,breaks=20,col="red",xlab="Mean values of step per day",main="Average Number of Steps Per Day\n NA removed")
abline(v=median(mean_steps$mean_of_steps),col="yellow",lwd=3)
```
### Average Daily Activity Pattern

Our **activity** dataset has 3 variables. Steps, date, interval. Interval is a variable that every value exists for every date. Just visualize:
````{r}
test_interval <- activity[activity$interval == 15,]
head(test_interval,10)
mean(test_interval$steps)
```
Example above shows us every interval observation value exist for all dates and steps vector so we need to iterate over all interval values corresponding steps vector. 

````{r}
mean_int_data <- aggregate(activity$steps, by=list(activity$interval),FUN=mean)
names(mean_int_data) <- c("interval","mean_of_steps")
head(mean_int_data,10)
with(mean_int_data,plot(interval,mean_of_steps,type="l",col="red",xlab="Interval",ylab="Average Steps",lwd=3,main="Time series of average 5-minute interval of average steps per day"))
```

Interval that has maximum steps:
```{r}
max_interval <- mean_int_data$interval[mean_int_data$mean_of_steps == max(mean_int_data$mean_of_steps)]
max_interval
````

###Imputing Missing Values

So far we worked with dataset called **activity**. But we cleared all the __NA__ values from original data set. So we are swicthing back to **activity12**. First of all we know that %13 percent of the data was missing. But we need to find exact number of missing rows. 
````{r}
total_NA <- sum(is.na(activity12$steps))
```
Total number of missing value is **`r total_NA`.** 

For replacing missing values with test values we need to define new observation values for steps variable. 
````{r}
# Which observations are missing
row_NA <- which(is.na(activity12$steps))

new_value <- rep(mean(activity12$steps,na.rm=TRUE),length(row_NA))

head(row_NA)
head(new_value)
````
So we will replace __NA__ values with **`r new_value[1]`.** row_NA is a vector that shows missing value positions.
```{r}
activity12[row_NA,"steps"] <- new_value
head(activity12,15)
sum(is.na(activity12))
str(activity12)
```
So we can follow up the same steps we did above with aggregate() function and hist().
````{r}
total_steps12 <- aggregate(activity12$steps,by=list(activity12$date),FUN=sum)
## it is nice to set col names after aggregation. 
names(total_steps12) <- c("date","sum_of_steps")
hist(total_steps12$sum_of_steps,breaks=20,col="blue",xlab="Total Steps",main="Total Number of Steps Per Day\n NA filled with mean value")
mean12 <- mean(total_steps12$sum_of_steps)
median12 <- mean(total_steps12$sum_of_steps)
abline(v=mean12,lwd=7,col="yellow")
abline(v=median12,lwd=3,col="red")
mean12
median12
```

###Difference Betweem Weekdays and Weekends

We are going to use weekdays() function to add weekdays based on date variable. 

```{r}
activity12$weekday <- weekdays(as.Date(activity12$date))
activity12$weekday.type <- factor(ifelse(activity12$weekday == "Sunday" | activity12$weekday == "Saturday","weekend","weekday"),levels=c("weekday","weekend"))
head(activity12,10)
tail(activity12,10)
```

We created a new dataset now we are going to make a panel plot for it.
```{r}
# we need to calculate average of steps per weekday
mean_act12 <- aggregate(activity12$steps,by=list(activity12$weekday.type,activity12$weekday, activity12$interval), FUN=mean)
names(mean_act12) <- c("type","weekday","interval","mean_of_steps")
head(mean_act12)


library(ggplot2)
library(lattice)
xyplot(mean_of_steps ~ interval | type,data=mean_act12,type ="l",layout=c(1,2),col="red",ylab="Steps",xlab="Interval",main="Weekdays vs. Weekends")


```

